---
title: "Practice Lecture 8 MATH 390.4 Queens College"
author: "Professor Adam Kapelner"
date: "March 26, 2019"
---


# The Grammar of graphics and ggplot

First load the package and the dataset of interest as a dataframe:

```{r}
pacman::p_load(ggplot2, quantreg)
cars = MASS::Cars93 #dataframe
```

ggplot is based on the "Grammar of Graphics", a concept invented by the Statistician / Computer Scientist Leland Wilkinson who worked on SPSS, Tableau and now he works at H20, software that analyzes big data. The reference of interest is [here](http://papers.rgrossman.com/proc-094.pdf). He drew on ideas from John Tukey (one of the great statistician of the previous generation) while he was at Bell Labs, Andreas Buja (one of my professors at Penn) and Jerome Friedman (the professor that taught my data mining course at Stanford - remember the diamonds story?) 

It is a language that allows us to describe the components of a graphic. Previously, graphics were done in one shot and it was clunky. ggplot is a library written by Hadley Wickham based on this concept. Wickham is probably the most famous person in statistical computing today. He has commit rights in R and is one of the architects of RStudio. He calls grammar of graphics "graphical poems". Here are the basic components:

* an underlying data frame
* an "aesthetic" that maps visualization axes in the eventual plot(s) to variables in the data frame
* a "layer" which is composed of
  - a geometric object
  - a statistical transformation
  - a position adjustment
* a "scale" for each aesthetic
* a "coordinate" system for each aesthetic
* optional "facets" (more graphics)
* optional "labels" for the title, axes title, points, etc.

Don't worry - everything has "smart defaults" in Wickham's implementation so you don't have to worry about most things. We will explore some of the features below. Here's a good [cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf).

ggplot is layered where each component is an object. The objects are added to each other since the "+" operator is overloaded to accept these additions. This is nice because each component can be saved and reused. The following initialized the graphics:

```{r}
ggplot(cars)
```

Nothing happened - except the plot window went gray (the smart default). This is already rendering a graphic, but since it hasn't been given any of the required information, it has nothing to display. Next we create an aesthetics indicating a one-way plot (one variable only).

```{r}
ggplot(cars) + 
  aes(Price)
```

Notice how it can understand the variable name as an object name.

Since we've given it an aesthetics object, it now knows which variable is the x axis (default). It already knows the ranges of the variable (a smart default) and a default scale and coordinate system (smart defaults).

Usually this is done in one step by passing the aesthetics object into the ggplot:

```{r}
ggplot(cars, aes(Price))
```

Now we need to pick a layer by specifying a geometry. This is a type of plot. Since the predictor type of price is continuous, let's pick the "histogram" using the `geom_histogram` function:

```{r}
ggplot(cars, aes(Price)) +
  geom_histogram()
```

This can be customized:

```{r}
ggplot(cars, aes(Price)) +
  geom_histogram(binwidth = 1, col = "darkgreen", fill = "blue", alpha = 0.4)
```

Want to save it for your latex?

```{r}
ggsave("plot.png")
system("open plot.png")
ggsave("plot.pdf")
system("open plot.pdf")
```

Here are some other options besides the histogram:

```{r}
ggplot(cars, aes(Price)) +
  geom_dotplot()
ggplot(cars, aes(Price)) +
  geom_area(stat = "bin", binwidth = 2)
ggplot(cars, aes(Price)) +
  geom_freqpoly()
ggplot(cars, aes(Price)) +
  geom_density(fill = "green", alpha = 0.4)
```


Can we compare price based on different conditions? Yes, we can subset the data and use color and alpha:

```{r}
ggplot(cars, aes(Price)) +
  geom_density(data = subset(cars, Man.trans.avail == "Yes"), col = "grey", fill = "darkgreen", alpha = 0.4) +
  geom_density(data = subset(cars, Man.trans.avail == "No"), col = "grey", fill = "red", alpha = 0.4)
```

Sidebar: why are cars that have manual transmissions available cheaper?

We can look at this also using a smoothed estimate of the conditional distributions:

```{r}
ggplot(cars, aes(Price)) +
  geom_histogram(data = subset(cars, Man.trans.avail == "Yes"), binwidth = 1, col = "grey", fill = "darkgreen", alpha = 0.4) +
  geom_histogram(data = subset(cars, Man.trans.avail == "No"), binwidth = 1, col = "grey", fill = "red", alpha = 0.4)
```

What if the variable is not continuous e.g. Cylinders? We can use a bar graph / bar plot.

```{r}
ggplot(cars, aes(Cylinders)) +
  geom_bar()
```

This is essential frequency by level of the categorical variable.

Now let's move on to looking at one variable versus another variable. For example price by engine power:

```{r}
ggplot(cars, aes(x = Horsepower, y = Price))
```

Since we've given it an aesthetics object, it now knows which variable is the x axis and which variable is the y axis. It already knows the ranges of the variables (a smart default) and a default scale and coordinate system (smart defaults).

Just as before, now we need to pick a layer by specifying a geometry. This is a type of plot. Let's pick the "scatterplot" using the `geom_point` function:

```{r}
ggplot(cars, aes(x = Horsepower, y = Price)) +
  geom_point()
```

Now we have a nice scatterplot. This function uses the inherited data, the inherited aesthetics. Since this "geometry" is a "layer", we can pass in options to the layer.

```{r}
base_and_aesthetics = ggplot(cars, aes(x = Horsepower, y = Price))
base_and_aesthetics + geom_point(col = "red", fill = "green", shape = 23, size = 3, alpha = 0.3)
```

Let's handle names of axes, title and ranges:

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics +
  ggtitle("Average Car Price vs. Engine Power", subtitle = "in the Cars93 dataset") +
  ylab("Price (in $1000's)")
base_and_aesthetics_with_titles +
  xlim(0, 400) +
  ylim(0, 50) +
  geom_point()
```

Let's transform the variables:

```{r}
base_and_aesthetics_with_titles +
  scale_x_continuous(trans = "log2") +
  geom_point()
```

Each unit increase on the x-axis now represent a doubling increase in x (although the whole scale only spans 3 units). But look at how the grid didn't keep up. Let's fix this:

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  scale_x_continuous(trans = "log2", breaks = round(seq(0, max(cars$Horsepower), length.out = 6)))
```

We can do the same to the y axis:

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  scale_y_continuous(trans = "log2")
  scale_x_continuous(trans = "log2", breaks = round(seq(0, max(cars$Horsepower), length.out = 6)))
```

Let's look at some more geometries.

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  geom_smooth()
```

Here, I've added two geometries on the same aesthetic! This attempts to explain the relationship $f(x)$ using smoothing. Let's go for more.

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  geom_smooth() +
  geom_rug()
```

This allows us to also see the marginal distributions of the two variables.

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  geom_smooth() +
  geom_quantile(col = "red") +
  geom_rug()
```

This fits a line and tries to indicate statistical significance of the line. We have *not* covered any statistics in this class yet (ironic!) ... so ignore how the window is generated.

Can we display more than two dimensions? Yes. Let's indicate a third dimension with shape (only works with factors).

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics_with_titles +
  ggtitle("Average Car Price by Power and Transmission")
base_and_aesthetics_with_titles +
  geom_point(aes(shape = Man.trans.avail)) +
  geom_smooth() +
  geom_rug()
```

Can we display more than three dimensions? Yes. Let's indicate a fourth dimension with color.

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics_with_titles +
  ggtitle("Average Car Price by Power, Transmission & Drivetrain")
base_and_aesthetics_with_titles +
  geom_point(aes(shape = Man.trans.avail, col = DriveTrain)) +
  geom_smooth() +
  geom_rug()
```

Can we go to a fifth dimension? Maybe?

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics_with_titles +
  ggtitle("Average Car Price by Power, Transmission & Drivetrain")
base_and_aesthetics_with_titles +
  geom_point(aes(shape = Man.trans.avail, col = DriveTrain, alpha = Weight)) + #size?
  geom_smooth() +
  geom_rug()
```

A seventh? We can use text labels adjacent to the scatterplot's points.

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics_with_titles +
  ggtitle("Average Car Price by Power, Transmission, Drivetrain,  Weight & #Cylinders")
base_and_aesthetics_with_titles +
  geom_point(aes(shape = Man.trans.avail, col = DriveTrain, alpha = Weight)) + #size?
  geom_text(aes(label = Cylinders), vjust = 1.5, col = "darkgrey", lineheight = 0.3, size = 3) +
  geom_smooth() +
  geom_rug()
```

Getting difficult to see what's going on.

Let's move away from the scatterplot to just density estimation:

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics_with_titles +
  ggtitle("Average Car Price by Power") #reset the title
base_and_aesthetics_with_titles +
  geom_density2d()
```

Other alternatives:

```{r}
base_and_aesthetics_with_titles +
  geom_bin2d(binwidth = c(8, 3))
pacman::p_load(hexbin)
base_and_aesthetics_with_titles +
  geom_hex()
```

This is like a two-dimensional histogram where the bar / hexagon heights are seen with color.

What if the x-axis is categorical for example Cylinders versus price? Typical is the "box and whiskers" plot:

```{r}
ggplot(cars, aes(x = Cylinders, y = Price)) +
  geom_boxplot()
```

Clear relationship!

How about multiple subplots based on the subsetting we did in the histograms? This is called "faceting". Here are two bivariate visualizations laid horizontally:

```{r}
ggplot(cars, aes(x = Horsepower, y = Price)) +
  geom_point() +
  geom_smooth() +
  facet_grid(. ~ Man.trans.avail)
```

Or alternatively, vertically:

```{r}
ggplot(cars, aes(x = Horsepower, y = Price)) +
  geom_point() +
  geom_smooth() +
  facet_grid(Man.trans.avail ~ .)
```

And we can even double-subset:

```{r}
ggplot(cars, aes(x = Horsepower, y = Price)) +
  geom_point() +
  facet_grid(Man.trans.avail ~ Origin, scales = "free")
```

And we can even triple-subset or more:

```{r}
cars$MedWeight = ifelse(cars$Weight > median(cars$Weight), ">MedWeight", "<MedWeight")
ggplot(cars, aes(x = Horsepower, y = Price)) +
  geom_point() +
  facet_grid(Man.trans.avail ~ Origin + MedWeight, scales = "free")
```

These three varibles seem somewhat independent.

There are other primitives like `geom_abline` which graphs a line and `geom_segment` we will see today. Note that if you want plots rendered within functions or loops you have to explicitly call the `plot` function:

```{r}
for (nsim in 1 : 3){
  graphics_obj = ggplot(data.frame(x = rnorm(1000))) + geom_histogram(aes(x))
  graphics_obj
}
```

versus:

```{r}
for (nsim in 1 : 3){
  graphics_obj = ggplot(data.frame(x = rnorm(1000))) + geom_histogram(aes(x))
  plot(graphics_obj)
}
```


Lastly, ggplot offers lots of nice customization themes:

```{r}
graphics_obj = base_and_aesthetics_with_titles +
  geom_point() +
  geom_smooth() +
  geom_quantile(col = "red") +
  geom_rug()
graphics_obj + theme_bw()
graphics_obj + theme_dark()
graphics_obj + theme_classic()

```

Packages offer even more:

```{r}
pacman::p_load(forcats, lazyeval, ggthemes)
graphics_obj + theme_economist()
graphics_obj + theme_stata()
graphics_obj + theme_tufte()
```

and of course, the whimsical one and only:


```{r}
pacman::p_load(xkcd, extrafont)
download.file("http://simonsoftware.se/other/xkcd.ttf", dest = "xkcd.ttf", mode = "wb")
#MAC
# system("mv xkcd.ttf /Library/Fonts")
# font_import(path = "/Library/Fonts", pattern = "xkcd", prompt = FALSE)
# fonts()
# fonttable()
# loadfonts()
#WINDOWS
font_import(path = ".", pattern = "xkcd", prompt = FALSE)
fonts()
fonttable()
loadfonts(device="win")

graphics_obj + theme_xkcd()
```

# C++ and R

R goes back to 1995 when it was adapted from S (written in 1976 by John Chambers at Bell Labs) with minor modifications. The core of base R is written in C and Fortran. These two languages are the fastest known languages (how to measure "fastest" is a huge debate). Thus, base R is very fast. For instance the `sort` function is as fast as C/Fortran since it immediately calls compiled C/Fortran routines.

However, R code itself that you write is "interpreted" which means it is not compiled until you run it. And it has to compile on-the-fly, making it very slow. Prior to v3.4 (April, 2017) it was even slower since the code wasn't JIT compiled. All this "real CS" stuff you can learn in another class..

One notable place to observe this slowness relative to other languages is in looping. For example:


```{r}
f = function(){
  SIZE = 1e7
  v = array(NA, SIZE)
  for (i in 1 : SIZE){
    v[i] = i
  }
  v
}

v = f()
```

How long does this take?

```{r}
# install.packages("Rcpp")
# pacman::p_load_current_gh("hadley/lineprof")
# lineprof(f())

system.time({
  f()
})
#on my office computer for SIZE = 10,000,000:
# user  system elapsed 
# 7.13    0.20    7.35 
```

Take a simple function that computes square roots on each element:

```{r}
sqrt_vector = function(v){
  v_new = array(NA, length(v))
  for (i in 1 : length(v)){
    v_new[i] = sqrt(v[i])
  }
  v_new
}

system.time({
  sqrt_vector(v)
})
```

Does the apply function help?

```{r}
system.time({
  apply(v, MARGIN = 1, FUN = sqrt)
})
```

Strange that this takes so long? So it doesn't help... it hurts.

How much faster in C++ should this be?

Enter the `Rcpp` package - a way to compile little bits of C++ on the fly.

```{r}
pacman::p_load(Rcpp)
```

Let's write this for loop function to sqrt-ize and compile it and then save it into our namespace to be called like a regular function.

```{r}
cppFunction('
  NumericVector sqrt_vector_cpp(NumericVector v) {
    int n = v.size();
    NumericVector v_new(n);
    for (int i = 0; i < n; i++) { //indices from 0...n-1 not 1...n!
      v_new[i] = sqrt(v[i]);
    }
    return v_new;
  }
')
```

What do these two functions look like?

```{r}
sqrt_vector
sqrt_vector_cpp
```

One shows the R code and then says it is bytecode-compiled (go to an advanced CS class). The other just says we `.Call` some C++ function in a certain address and the argument to be inputted.

What is the gain in runtime?

```{r}
system.time({
  sqrt_vector_cpp(v)
})
```

WOW. 10x!!! Can't beat that with a stick...

Let's do a not-so-contrived example...

Matrix distance... Let's compute the distances of all pairs of rows in a dataset.

```{r}
X_boston = as.matrix(MASS::Boston[, 1 : 13])

compute_distance_matrix = function(X){
  n = nrow(X)
  D = matrix(NA, n, n)
  for (i_1 in 1 : (n - 1)){
    for (i_2 in (i_1 + 1) : n){
      D[i_1, i_2] = sqrt(sum((X[i_1, ] - X[i_2, ])^2))
    }
  }
  D
}

cppFunction('
  NumericMatrix compute_distance_matrix_cpp(NumericMatrix X) {
    int n = X.nrow();
    int p = X.ncol();
    NumericMatrix D(n, n);
    std::fill(D.begin(), D.end(), NA_REAL);

    for (int i_1 = 0; i_1 < (n - 1); i_1++){
      for (int i_2 = i_1 + 1; i_2 < n; i_2++){
        int sqd_diff = 0;
        for (int j = 0; j < p; j++){
          sqd_diff += pow(X(i_1, j) - X(i_2, j), 2); //by default the cmath library in std is loaded
        }
        D(i_1, i_2) = sqrt(sqd_diff); //by default the cmath library in std is loaded
      }
    }
    return D;
  }
')

```

```{r}
system.time({
  D = compute_distance_matrix(X_boston)
})
round(D[1 : 5, 1 : 5], 2)
```

Slow...

```{r}
system.time({
  D = compute_distance_matrix_cpp(X_boston)
})
round(D[1 : 5, 1 : 5], 2)
```

Lightning...

Writing functions as strings that compile is annoying. It is better to have separate files. For instance...

```{r}
sourceCpp("distance_matrix.cpp")
```

Another place where C++ pays the rent is recursion. Here is a quicksort implementation in R taken from 

```{r}
quickSort <- function(arr) {
  # Pick a number at random.
  mid <- sample(arr, 1)

  # Place-holders for left and right values.
  left <- c()
  right <- c()
  
  # Move all the smaller values to the left, bigger values to the right.
  lapply(arr[arr != mid], function(d) {
    if (d < mid) {
      left <<- c(left, d)
    }
    else {
      right <<- c(right, d)
    }
  })
  
  if (length(left) > 1) {
    left <- quickSort(left)
  }
  
  if (length(right) > 1) {
    right <- quickSort(right)
  }
  
  # Finally, return the sorted values.
  c(left, mid, right)
}
```

Let's make sure it works and find the time it takes to run:

```{r}
n = 5e4
x = rnorm(n)
system.time({
  x_sorted_ours = quickSort(x)
})

```

Let's make sure it works and profile R's `sort` function.

```{r}
system.time({
  x_sorted_R = sort(x)
})
pacman::p_load(testthat)
expect_equal(x_sorted_ours, x_sorted_R)
```

Basically infinitely faster. Let's make our own C++ implementation.



