---
title: "Practice Lecture 9 MATH 390.4 Queens College"
author: "Professor Adam Kapelner"
date: "April 2, 2019"
---

## Dplyr

Now that we know piping, we can start adding some nice functions that manipulate data frames. Let's look at the diamonds dataset and load the dplyr library.

```{r}
data(diamonds, package = "ggplot2")
pacman::p_load(dplyr)
```

Let's remind ourselves of the dataset:

```{r}
str(diamonds)
summary(diamonds)
```

The package `dplyr` offers many conveninent functions to manipulate, clean, put together data frames (AKA "munging", "wrangling"). It works really nicely with the piping chain as you "begin" the manipulation with the dataset and then iteratively do step 1, step 2, etc until you wind up with what end product you would like.

Beginning with the most obvious, `rename` will rename a column

```{r}
diamonds %>% 
  rename(weight = carat)
```


The `select` function selects columns in the order you ask it to.

```{r}
diamonds %>% 
  select(cut, carat, price) #these three only in this order
diamonds %>% 
  select(carat, price, cut) #these three only
diamonds %>% 
  select(-c(x, y, z)) #leave out these three
diamonds %>% 
  select(-x, -y, -z) #leave out these three
```

If you want to rearrange the columns, you pretend to select a subset and then ask for everything else:

```{r}
diamonds %>% 
  select(carat, price, cut, everything()) #these three in this order first then everything else
diamonds %>% 
  select(-carat, everything()) #move carat last (first drop it, and then add it back in with everything)
```


The `arrange` method sorts the rows:

```{r}
diamonds %>%
  arrange(carat) #default is ascending i.e. lowest first
diamonds %>%
  arrange(desc(carat)) #switch to descending, i.e. highest first
diamonds %>%
  arrange(desc(color), desc(clarity), desc(cut), desc(carat)) #multiple sorts - very powerful
```



The filter method subsets the data based on conditions:

```{r}
diamonds %>%
  filter(cut == "Ideal")
diamonds %>%
  filter(cut == "Ideal") %>%
  filter(depth < 65)
diamonds %>%
  filter(cut == "Ideal") %>%
  filter(depth < 65) %>%
  filter(x * y * z > 20)
diamonds %>%
  filter(cut == "Ideal" & depth < 65 & x * y * z > 20)
diamonds %>%
  filter((cut == "Ideal" | cut == "Premium") & depth < 65 & x * y * z > 20)
diamonds %>%
  filter(cut %in% c("Ideal", "Premium") & depth < 65 & x * y * z > 20)
```

How about removing all rows that are the same?

```{r}
diamonds
diamonds %>%
  distinct
unique(diamonds$carat) #there's only a few weight measurements that are possible...
diamonds %>%
  distinct(carat, .keep_all = TRUE) #keeps the first row for each unique weight measurement
```

Sampling is easy

```{r}
diamonds %>%
  sample_n(20)
0.0005 * nrow(diamonds)
diamonds %>%
  sample_frac(0.0005)
diamonds %>%
  slice(5000 : 5009)
```

There are many ways to reshape a dataset. We will see two now and a few functions later when it becomes important. For instance: we can collapse columns together using the `unite` function from package `tidyr`.

```{r}
pacman::p_load(tidyr)
diamonds2 = diamonds %>%
  unite(dimensions, x, y, z, sep = " x ")
diamonds2
```

We can reverse this operation by separating them out using `separate`:

```{r}
diamonds2 %>%
  separate(dimensions, c("x", "y", "z"), sep = " x ")
```

Now for some real fun stuff. Let's create new features with the `mutate` function.

```{r}
diamonds %>%
  mutate(volume = x * y * z) #adds a new column keeping the old ones (this was our exam problem)
diamonds %>%
  mutate(price_per_carat = price / carat) %>%
  arrange(desc(price_per_carat))
```

Or rewrite old ones.

```{r}
diamonds %>%
  mutate(cut = substr(cut, 1, 1))
diamonds %>%
  mutate(carat = factor(carat))
```

Here are some more ways to create new variables:

```{r}
diamonds %>%
  mutate(carat = factor(ntile(carat, 5)))
diamonds %>%
  mutate(carat = percent_rank(carat))
diamonds %>%
  mutate(lag_price = lag(price)) #if this data was a time series
diamonds %>%
  mutate(cumul_price = cumsum(price)) #%>% tail
```

There are tons of package to do clever things. For instance, here's one that does dummies:

```{r}
pacman::p_load(sjmisc, snakecase)
diamonds %>%
  to_dummy(color, suffix = "label") %>% #this creates all the dummies
  bind_cols(diamonds) %>% #now we have to add all the data back in
  select(-matches("_"), everything()) %>% #this puts the dummies last
  select(-color) #finally we can drop color
diamonds %>% #convert all to dummies
  select(color, cut, clarity) %>%
  to_dummy(suffix = "label")
```


What if you want to create a new variable based on functions only run on subsets of the data. This is called "grouping".

For instance:

```{r}
diamonds %>%
  group_by(color) #nothing happened... this just is a directive to dplyr to do things a bit differently now
diamonds %>%
  group_by(color) %>%
  mutate(price_rank_within_color = dense_rank(price)) #creates a new feature based on running the feature only within group
diamonds %>%
  group_by(color) %>%
  mutate(avg_price_for_color = mean(price)) #creates a new feature based on running the feature only within group
```

How do you summarize data within group?

```{r}
diamonds %>%
  group_by(color) %>%
  summarize(avg_price = mean(price)) #where did all the other columns go???
diamonds %>%
  group_by(color) %>%
  summarize(avg_price = mean(price), sd_price = sd(price), count = n()) #where did all the other columns go???
diamonds %>%
  group_by(color) %>%
  summarize(avg_price = mean(price), avg_carat = mean(carat)) #where did all the other columns go???
diamonds %>%
  group_by(color) %>%
  summarize(min_price = min(price), med_price = median(price), max_price = max(price))
```

Putting it all together: ususally you're doing this manipulation get the dataset you want. Usually you're editing the dataset for real. Let's make a copy first:

```{r}
diamonds2 = diamonds
```

We first note that if we want to overwrite we can do:

```{r}
diamonds2 = diamonds2 %>%
  select(-x, -y, -z) %>%
  filter(carat < 0.5)
```

Or we can use a mutate operation that reads and writes simultaneously:

```{r}
pacman::p_load(magrittr)
diamonds2 = diamonds
diamonds2 %<>%
  select(-x, -y, -z) %>%
  filter(carat < 0.5) %>%
  arrange(carat, cut, color)
diamonds2
```

This is as far as we can go with dplyr right now given that this dataset doesn't have datetime information, some duplication among rows, missing data and given that there's not multiple dataframes. We will return to dplyr under these situations in the future.

