---
title: "Practice Lecture 2 MATH 390.4 Queens College"
author: "Professor Adam Kapelner"
date: "February 7, 2019"
---

## Programming! Conditional statements, loops, functions

We will now learn how to program in R.

* Conditional Statements (and ways to print to screen too).

```{r}
x = 3
if (x > 0){
  cat("The value of x is", x, "\n") #concatenate and print to screen
} else {
  print("The value of x is"); print(x) #each gets its own line
}

if (x > 0){
  ##
} 
else { ##NOTE: ELSE MUST GO ON SAME LINE AS THE IF'S CLOSE BRACE!!! This is not Java or C++!
  ##
}

x = -5
if (x > 0){
  cat("The value of x is", x, "\n") #concatenate and print to screen
} else if (x < -3) { ##NOTE: ELSE MUST GO ON SAME LINE AS THE IF'S CLOSE BRACE!!!
  print(paste("The value of x is", x, collapse = " ")) #paste with collapse is like a string join
} else {
  print("The value of x is"); print(x) #each gets its own line
}
```

There is a one line convenience function called ifelse:

```{r}
x = 3
a = ifelse(x > 0, 1, -1)
a
x = -5 : 5
a = ifelse(x > 0, 1, -1)
a
```



There is also a switch in R. It is _not_ an official keyword (despite RStudio's endorsement), but it is useful for non-vector character expressions:

```{r}
ch = "one"
# ch = c("one", "two")
switch(ch,
   "one" = 1,
   "two" = 2,
   "three" = 3,
   -99
)
```


* Looping

A for loop contains three ingredients: (a) a looping variable joined with keyword "in" to a (b) vector or list (which we will discuss later) and executes over an (c) expression recommended to be enclosed in curly braces:

```{r}
for (i in 1 : 10){
  j = i
  print(j)
}
```

Note that the looping variable i still exists but the variable within the looped expression j does not:

```{r}
i
j
```

You can iterate over a vector as well:

```{r}
vec = seq(1, 10)
for (i in vec){
  print(i)
}
```

And the vector does not have to be numeric:

```{r}
vec = c("a", "b", "c")
for (ch in vec){
  print(ch)
}
```

There is also the while loop but not a "do-while" loop

```{r}
x = 1
while (x <= 10){
  print(x)
  x = x + 1 #note: there is no "++" incrementing in R
}
```

This can also be done using a break:

```{r}
x = 1
while (TRUE){
  print(x)
  x = x + 1 #note: there is no "++" incrementing in R
  if (x > 10){
    break
  }
}
```

The "repeat" keyword is the same as the while-true:

```{r}
x = 1
repeat {
  print(x)
  x = x + 1 #note: there is no "++" incrementing in R
  if (x > 10){
    break
  }
}
```

The final control keyword tells the loop to break the current iteration but continue to the next:

```{r}
xs = rep(NA, 10)
xs[3] = 3
xs[7] = 7
xs

tot = 0
for (x in xs){
  if (is.na(x)){
    next #move to next element in the list in the for loop
  }
  tot = tot + x
}
tot
```

Note: "real" R programmers avoid loops since they are slow. Instead they use the "apply" function (but not "sapply" or "lapply" which are not faster). We will cover these in the future. The latest trend is to use in-line C++ code (we probably won't get to this).

* Functions

```{r}
my_function = function(x){
  x
}
##You may be used to:
#int my_function(int x){ 
#  return x;
#}
```

* Functions are objects in R. This is why you are actually assigning the function to an object.
* We don't need to declare type since R is not "statically typed" (higher level languages usually are not statically typed). Objects can be coerced into different types on the fly (R is "dynamically typed").
* No need for a "return" statement as the last line is the data that is returned. It is considered bad style to use "return" in R.

Let's make sure this works

```{r}
my_function(3)
my_function(x = 3) #you can specify that 3 is the value for argument "x"
my_function(y = 3) #illegal argument
```

R is somewhat user friendly as it allows for default argument values, making those arguments optional to specify when calling the function:

```{r}
my_function = function(x = 1, y = 2, z = 3, p = 4, q = 5, r = 6){
  (x + y + z) / (p + q + r)
}
my_function() #default execution
my_function(p = 0) #one optional argument specified, others defaulted
my_function(y = -2, q = 0) #two optional arguments specified, others defaulted

my_function = function(x, y = 2, z = 3, p = 4, q = 5, r = 6){
  (x + y + z) / (p + q + r)
}
my_function() #no dice
my_function(0) #required argument specified
my_function(0, q = 7) #required argument specified and one optional argument
rm(my_function) #can be deleted since it's an object
```


Now back to the in-class example...

## First modeling exercise

Before we model, let's fabricate the training data! Let's try to build a data matrix similar to the one in the class example. Let's imagine $n = 100$ and $x_1$ is salary, $x_2$ is a dummy variable for missing loan payment in their credit history, $x_3$ is a ordinal variable for crime type coded 0, 1, 2 or 3.

We can "make up" a dataset using the sampling we just learned.

```{r}
n = 100 #number of historical objects: the people
p = 3 #number of features about each

X = matrix(NA, nrow = n, ncol = p)
X
```

Some more useful matrix functions:

```{r}
nrow(X)
ncol(X)
dim(X)
length(X) #countless bugs!
c(X) #vectorizes / flattens a matrix (sometimes useful)
```

Why should we fill up this matrix with NA's? No technical reason; it is done for a practical reason. Every value is currently "missing" until it's filled in i.e. it will let you know if you didn't fill any of the values.

We can also name rows and columns. Each row is a historical person and each column is a feature about that person.

```{r}
colnames(X) = c(
  "salary", 
  "has_past_unpaid_loan", 
  "past_crime_severity"
)
colnames(X) #setter and a getter
fake_first_names = c(
  "Sophia", "Emma", "Olivia", "Ava", "Mia", "Isabella", "Riley", 
  "Aria", "Zoe", "Charlotte", "Lily", "Layla", "Amelia", "Emily", 
  "Madelyn", "Aubrey", "Adalyn", "Madison", "Chloe", "Harper", 
  "Abigail", "Aaliyah", "Avery", "Evelyn", "Kaylee", "Ella", "Ellie", 
  "Scarlett", "Arianna", "Hailey", "Nora", "Addison", "Brooklyn", 
  "Hannah", "Mila", "Leah", "Elizabeth", "Sarah", "Eliana", "Mackenzie", 
  "Peyton", "Maria", "Grace", "Adeline", "Elena", "Anna", "Victoria", 
  "Camilla", "Lillian", "Natalie", "Jackson", "Aiden", "Lucas", 
  "Liam", "Noah", "Ethan", "Mason", "Caden", "Oliver", "Elijah", 
  "Grayson", "Jacob", "Michael", "Benjamin", "Carter", "James", 
  "Jayden", "Logan", "Alexander", "Caleb", "Ryan", "Luke", "Daniel", 
  "Jack", "William", "Owen", "Gabriel", "Matthew", "Connor", "Jayce", 
  "Isaac", "Sebastian", "Henry", "Muhammad", "Cameron", "Wyatt", 
  "Dylan", "Nathan", "Nicholas", "Julian", "Eli", "Levi", "Isaiah", 
  "Landon", "David", "Christian", "Andrew", "Brayden", "John", 
  "Lincoln"
)
rownames(X) = fake_first_names
rownames(X) #setter and getter
X
```

Let's pretend "salary" is normally distributed with mean \$50,000 and standard error \$20,000. Let's make up some data

```{r}
X[, 1] = round(rnorm(n, 50000, 20000))
X
#another way to set this feature:
X[, "salary"] = round(rnorm(n, 50000, 20000))
X
```

A quick sidebar about vectors within matrices with row or column names:

```{r}
salaries = X[, 1] 
salaries #it's a vector with names
names(salaries) #access to its names
names(salaries)[3] = "Kapelner"
salaries
salaries[3]
salaries["Kapelner"]
sort(salaries)
#how do we sort salaries by name?
salaries[order(names(salaries))]
?order #it's like sort, but it returns indices
rm(salaries)
```

Are the salary values independent? Yes, or at least we assume so... Hopefully we will get to models in this semester where they are not independent.

We will eventually do visualization, but first let's take a look at a summary of this data:

```{r}
summary(X[, "salary"])
```

There are other base functions to know:

```{r}
mean(X[, "salary"]) #mean should be "average"!!
sum(X[, "salary"]) / length(X[, "salary"])
var(X[, "salary"])
sd(X[, "salary"])
median(X[, "salary"])
min(X[, "salary"])
max(X[, "salary"])
IQR(X[, "salary"])
```

There is also the convenient quantile and inverse quantile function

```{r}
quantile(X[, "salary"], probs = 0.5) 
quantile(X[, "salary"], probs = c(.1, .9))  
inverse_quantile_obj = ecdf(X[, "salary"]) #the "empirical" CDF
inverse_quantile_obj(50000)
inverse_quantile_obj(70000)
inverse_quantile_obj(0)
inverse_quantile_obj(-10000)
inverse_quantile_obj(200000)
```


Let's pretend "has_past_unpaid_loan" is benoulli distributed with probability 0.2

```{r}
X[, "has_past_unpaid_loan"] = rbinom(n, size = 1, prob = 0.2)
X
```

Is this a reasonable fabrication of this dataset? No... since salary and not paying back a loan are dependent r.v.'s. But... we will ignore this now.

It would be nice to see a summary of values. Would median and mean be appropriate here? No. For categorical variables, you should "table" them:

```{r}
table(X[, "has_past_unpaid_loan"])
```


Also, 50\% of people have no crime, 40\% have an infraction, 8\% a misdimeanor and 2\% a felony. Let's try to add this to the matrix. We first need to simulate this. Here's how:

```{r}
X[, "past_crime_severity"] = sample(
  c("no crime", "infraction", "misdimeanor", "felony"),
  size = n,
  replace = TRUE,
  prob = c(.50, .40, .08, .02)
)
X
```

Oh no - what happened?? Our matrix went all characters... The matrix type cannot handle numeric and categorical variables simultaneously! It would be nice to keep factor or character information in a matrix but this is not the spec.

Enter the key data type, the "data.frame" - this is the object that is used for modeling in the R ecosystem. It is essentially an upgraded matrix.

```{r}
X = data.frame(
  salary = round(rnorm(n, 50000, 20000)),
  has_past_unpaid_loan = rbinom(n, size = 1, prob = 0.2),
  past_crime_severity = sample(
    c("no crime", "infraction", "misdimeanor", "felony"),
    size = n,
    replace = TRUE,
    prob = c(.50, .40, .08, .02)
  )
)
row.names(X) = fake_first_names
X
```

RStudio gives us a nicer rendering of the information. You can open it up in a separate tab via:

```{r}
View(X)
```

and you can view summaries of each feature and data type of each feature via

```{r}
summary(X)
str(X)
```

Again, summary doesn't work for "has_past_unpaid_loan". We should convert it to factor and try again. Note the "$" operator which is now valid for data.frame objects.

```{r}
X$has_past_unpaid_loan = factor(X$has_past_unpaid_loan, labels = c("Never", ">=1"))
head(X) #make sure that worked
summary(X) #much better now!
str(X)
```

Now that we have two categorical variables, we can do a "cross tab":

```{r}
table(X$has_past_unpaid_loan)
table(X$past_crime_severity)
table(X$has_past_unpaid_loan, X$past_crime_severity)
#to avoid needing the "X$" over and over, use the convenience "with"
with(X,
  table(has_past_unpaid_loan, past_crime_severity)
)
```

In our training set D, we are missing one final variable, the response! Let's add it and say that 90\% of people are creditworthy i.e. they paid back their loan:

```{r}
X$paid_back_loan = factor(rbinom(n, size = 1, prob = 0.9), labels = c("No", "Yes"))
head(X) #make sure that worked
summary(X) #much better now!
```

Conceptually - why does this make no sense at all??? y is independent of X --- what happens then? No function f can ever have any predictive / explanatory power! This is just a silly example to show you the data types. We will work with real data soon. Don't worry.