---
title: "Practice Lecture 13 MATH 390.4 Queens College"
author: "Professor Adam Kapelner"
date: "May 7, 2019"
---

# Joins

```{r}
pacman::p_load(nycflights13, tidyverse, magrittr)
data(weather)
str(weather)
data(airports)
str(airports)
```

Note how the weather and airports datasets contain a common feature: name of airport. It is called `FAA` in airports and `origin` in weather.

First we rename the column in weather to match the column in airports:

```{r}
weather %<>% rename(faa = origin)
```

We also pare down the datasets so we can see the joins more clearly:

```{r}
airports %<>% select(faa, lat, lon)
weather %<>% select(faa, time_hour, temp, humid, wind_speed, pressure, wind_gust)
```

Some features just aren't measured that often e.g. `wind_gust`.

Let's do some joins. 

```{r}
head(airports)
head(weather)
airports_and_weather = left_join(airports, weather, by = "faa")
airports_and_weather %>% sample_n(500)
```


```{r}
head(airports)
head(weather)
airports_and_weather = right_join(airports, weather, by = "faa")
airports_and_weather %>% sample_n(500)
```

Just an example... drop `EWR` from airports:

```{r}
airports_without_EWR = airports %>%
  filter(faa != "EWR")
right_join(airports_without_EWR, weather, by = "faa") %>% sample_n(500)
```

```{r}
head(airports)
head(weather)
airports_and_weather = inner_join(airports, weather, by = "faa")
airports_and_weather %>% sample_n(500)
```




```{r}
head(airports)
head(weather)
airports_and_weather = full_join(airports, weather, by = "faa")
airports_and_weather %>% sample_n(500) %>% arrange(faa)
```


```{r}
airports_without_EWR = airports %>%
  filter(faa != "EWR")
full_join(airports_without_EWR, weather, by = "faa") %>% sample_n(500)
```

There is also `semi_join` and `anti_join` that do the opposite of joining. In my experience, these use cases are limited.

# Missingness

Take a look at the weather dataset again:

```{r}
str(weather)
summary(weather)
```

#Classification Trees and Confusion Tables

Let's load up the adult dataset where the response is 1 if the person makes more than $50K per year and 0 if they make less than $50K per year.

```{r}
pacman::p_load_gh("coatless/ucidata")
data(adult)
adult %<>% 
  na.omit #kill any observations with missingness
str(adult)
?adult
```

Let's use samples of 2,000 to run experiments:

```{r}
test_size = 2000
train_indices = sample(1 : nrow(adult), test_size)
adult_train = adult[train_indices, ]
y_train = adult_train$income
X_train = adult_train
X_train$income = NULL
test_indices = sample(setdiff(1 : nrow(adult), train_indices), test_size)
adult_test = adult[test_indices, ]
y_test = adult_test$income
X_test = adult_test
X_test$income = NULL
```

Load the tree package:

```{r}
options(java.parameters = "-Xmx4000m")
# pacman::p_install_gh("kapelner/YARF", subdir = "YARF", ref = "dev")
pacman::p_load(YARF)
```

Make a tree and compute in-sample and out of sample fits:

```{r}
tree_mod = YARFCART(X_train, y_train)
y_hat_train = predict(tree_mod, X_train)
y_hat_test = predict(tree_mod, X_test)
```

Let's look at the confusion table in-sample:

```{r}
table(y_train, y_hat_train)
```

There are no errors here! Thus, precision and recall are both 100%. This makes sense because classification trees overfit.

Let's do the same oos:

```{r}
oos_conf_table = table(y_test, y_hat_test)
oos_conf_table
```

We didn't do as well (of course). Let's calculate some performance metrics. We assume ">50k" is the "positive" category and "<=50k" is the "negative" category. Note that this choice is arbitrary and everything would just be switched if we did it the other way.

```{r}
n = sum(oos_conf_table)
n
fp = oos_conf_table[1, 2]
fn = oos_conf_table[2, 1]
tp = oos_conf_table[2, 2]
tn = oos_conf_table[1, 1]
num_pred_pos = sum(oos_conf_table[, 2])
num_pred_neg = sum(oos_conf_table[, 1])
num_pos = sum(oos_conf_table[2, ])
num_neg = sum(oos_conf_table[1, ])
acc = (tp + tn) / n
acc
misclassifcation_error = 1 - acc
misclassifcation_error
precision = tp / num_pred_pos
precision
recall = tp / num_pos
recall
false_discovery_rate = 1 - precision
false_discovery_rate
false_omission_rate = fn / num_pred_neg
false_omission_rate
```

Let's see how this works on a dataset whose goal is classification for more than 2 levels. Note: this is only possible now with trees!

```{r}
rm(list = ls())
pacman::p_load(mlbench)
data(LetterRecognition)
LetterRecognition = na.omit(LetterRecognition) #kill any observations with missingness
str(LetterRecognition)
?LetterRecognition
```

Now we split the data:

```{r}
test_samp = 2000
train_indices = sample(1 : nrow(LetterRecognition), test_samp)
ltr_train = LetterRecognition[train_indices, ]
y_train = ltr_train$lettr
X_train = ltr_train
X_train$lettr = NULL
test_indices = sample(setdiff(1 : nrow(LetterRecognition), train_indices), test_samp)
ltr_test = LetterRecognition[test_indices, ]
y_test = ltr_test$lettr
X_test = ltr_test
X_test$lettr = NULL
```

And fit a tree model and its in-sample and oos fits:

```{r}
tree_mod = YARFCART(X_train, y_train)
y_hat_train = predict(tree_mod, X_train)
y_hat_test = predict(tree_mod, X_test)
```

Take a look at the in-sample confusion matrix:

```{r}
table(y_train, y_hat_train)
```

Perfecto... as expected... 

Now the oos confusion matrix:

```{r}
oos_confusion_table = table(y_test, y_hat_test)
oos_confusion_table
```

Hard to read. Let's make it easier to read by blanking out the diagonal and looking at entried only >= 5:

```{r}
oos_confusion_table[oos_confusion_table < 5] = ""
diag(oos_confusion_table) = "."
oos_confusion_table
```

What's it using to determine letter?

```{r}
illustrate_trees(tree_mod, max_depth = 3, open_file = TRUE)
```

Where did these features comes from?? Deep learning helps to create the features from the raw pixel data. Wish I had a whole next semester to discuss this...


