---
title: "Practice Lecture 12 MATH 390.4 Queens College"
author: "Professor Adam Kapelner"
date: "April 30, 2019"
---


# Bias-Variance Decomposition of Generalization Error


Let's try to fit a quadratic $f$ with a linear model and examine bias-variance tradeoff.

```{r}
rm(list = ls())
xmin = 0
xmax = 5
n_train = 20
n_test = 1000
sigsq = 1
resolution = 10000

Nsim = 1000

training_gs = matrix(NA, nrow = Nsim, ncol = 2)
all_residuals = matrix(NA, nrow = Nsim, ncol = n_test)
x_trains = matrix(NA, nrow = Nsim, ncol = n_train)
y_trains = matrix(NA, nrow = Nsim, ncol = n_train)
for (nsim in 1 : Nsim){
  #simulate dataset $\mathbb{D}$
  x_train = runif(n_train, xmin, xmax)
  delta_train = rnorm(n_train, 0, sqrt(sigsq)) #mean zero, variance sigsq always (not dependent on value of x)
  y_train = x_train^2 + delta_train
  x_trains[nsim, ] = x_train
  y_trains[nsim, ] = y_train
  
  
  #fit a model g | x's, delta's and save it
  g_model = lm(y_train ~ ., data.frame(x = x_train))
  training_gs[nsim, ] = coef(g_model)
  
  #generate oos dataset and save residuals on oos data
  x_test = runif(n_test, xmin, xmax)
  delta_test = rnorm(n_test, 0, sigsq) #mean zero, variance sigsq always (not dependent on value of x)
  y_test = x_test^2 + delta_test
  y_hat_test = predict(g_model, data.frame(x = x_test))
  all_residuals[nsim, ] = y_test - y_hat_test
}
```

Graph a few:

```{r}
pacman::p_load(ggplot2)
ggplot() + 
  geom_point(aes(x, y), data = data.frame(x = x_trains[1, ], y = y_trains[1, ]), col = "blue") +
  geom_point(aes(x, y), data = data.frame(x = x_trains[2, ], y = y_trains[2, ]), col = "darkgreen") +
  geom_point(aes(x, y), data = data.frame(x = x_trains[3, ], y = y_trains[3, ]), col = "red")
```

The blue dataset is one possible $\mathbb{D}$, the green dataset is another possible $\mathbb{D}$ and the red dataset is another possible $\mathbb{D}$. This illustrated "dataset-dataset variability".

Take a look at the mse:

```{r}
mse = mean(c(all_residuals)^2)
mse
```

This should be irreducible error plus bias-squared plus variance.

Let's visualize the bias

```{r}
g_average = colMeans(training_gs)
f = function(x){x^2}
x = seq(xmin, xmax, length.out = resolution)
ggplot(data.frame(x = x, f = f(x)), aes(x, f)) + 
  geom_line(col = "darkgreen") + 
  geom_abline(intercept = g_average[1], slope = g_average[2], col = "red")
```

What is the average bias of $g$?

```{r}
x = seq(xmin, xmax, length.out = resolution)
g_avg_x = g_average[1] + g_average[2] * x
f_x = x^2
biases = f_x - g_avg_x
expe_bias_g_sq = mean(biases^2)
expe_bias_g_sq
```

What is the variance? Let's look at all lines:

```{r}
plot_obj = ggplot() + 
  xlim(xmin, xmax) + ylim(xmin^2, xmax^2)

for (nsim in 1 : min(Nsim, 100)){ #otherwise takes too long
  plot_obj = plot_obj + geom_abline(intercept = training_gs[nsim, 1], slope = training_gs[nsim, 2], col = "blue")
}

plot_obj +
  geom_abline(intercept = g_average[1], slope = g_average[2], col = "red", lwd = 2)
```

Now calculate this average variance:

```{r}
x = seq(xmin, xmax, length.out = resolution)

expe_g_x = g_average[1] + g_average[2] * x

var_x_s = array(NA, Nsim)
for (nsim in 1 : Nsim){
  g_x = training_gs[nsim, 1] + training_gs[nsim, 2] * x
  var_x_s[nsim] = mean((g_x - expe_g_x)^2)
}

expe_var_g = mean(var_x_s)
expe_var_g
```

Now check the equivalence

```{r}
mse
sigsq
expe_bias_g_sq
expe_var_g
sigsq + expe_bias_g_sq + expe_var_g
```

This is not exactly equal due to numerical error.

Let's try the whole thing again using a quadratic regression!

```{r}
training_gs = matrix(NA, nrow = Nsim, ncol = 3)
all_residuals = matrix(NA, nrow = Nsim, ncol = n_test)

for (nsim in 1 : Nsim){
  #simulate dataset $\mathbb{D}$
  x_train = runif(n_train, xmin, xmax)
  delta_train = rnorm(n_train, 0, sqrt(sigsq)) #mean zero, variance sigsq always (not dependent on value of x)
  y_train = x_train^2 + delta_train
  
  #fit a model g | x's, delta's and save it
  g_model = lm(y_train ~ poly(x, 2, raw = TRUE), data.frame(x = x_train))
  training_gs[nsim, ] = coef(g_model)
  
  #generate oos dataset and save residuals on oos data
  x_test = runif(n_test, xmin, xmax)
  delta_test = rnorm(n_test, 0, sigsq) #mean zero, variance sigsq always (not dependent on value of x)
  y_test = x_test^2 + delta_test
  y_hat_test = predict(g_model, data.frame(x = x_test))
  all_residuals[nsim, ] = y_test - y_hat_test
}
```

Take a look at the mse:

```{r}
mse = mean(c(all_residuals)^2)
mse
```

Much lower! Why? Bias went down. 

Let's visualize the bias

```{r}
g_average = colMeans(training_gs)
f = function(x){x^2}
x = seq(xmin, xmax, length.out = resolution)
ggplot(data.frame(x = x, f = f(x)), aes(x, f)) + 
  geom_line(col = "darkgreen") + 
  stat_function(fun = function(x){g_average[1] + g_average[2] * x + g_average[3] * x^2}, col = "red")
```

Not much! What is the average bias of $g$?

```{r}
x = seq(xmin, xmax, length.out = resolution)
g_avg_x = g_average[1] + g_average[2] * x + g_average[3] * x^2
f_x = x^2
biases = f_x - g_avg_x
expe_bias_g_sq = mean(biases^2)
expe_bias_g_sq
```

What is the variance? Let's look at all lines:

```{r}
plot_obj = ggplot(data.frame(x = x)) + 
  xlim(xmin, xmax) + ylim(xmin^2, xmax^2)

for (nsim in 1 : min(Nsim, 50)){ #otherwise takes too long
  plot_obj = plot_obj + geom_line(data = data.frame(x = x, y = training_gs[nsim, 1] + training_gs[nsim, 2] * x + training_gs[nsim, 3] * x^2), mapping = aes(x, y), col = "blue")
}

plot_obj +
  stat_function(fun = function(x){g_average[1] + g_average[2] * x + g_average[3] * x^2}, col = "red", lwd = 2)
```

Now calculate this average variance:

```{r}
x = seq(xmin, xmax, length.out = resolution)

expe_g_x = g_average[1] + g_average[2] * x + g_average[3] * x^2

var_x_s = array(NA, Nsim)
for (nsim in 1 : Nsim){
  g_x = training_gs[nsim, 1] + training_gs[nsim, 2] * x + training_gs[nsim, 3] * x^2
  var_x_s[nsim] = mean((g_x - expe_g_x)^2)
}

expe_var_g = mean(var_x_s)
expe_var_g
```

Now check the equivalence

```{r}
mse
sigsq
expe_bias_g_sq
expe_var_g
sigsq + expe_bias_g_sq + expe_var_g
```


Try it again with quintic polynomials!

```{r}
training_gs = matrix(NA, nrow = Nsim, ncol = 6)
all_residuals = matrix(NA, nrow = Nsim, ncol = n_test)

for (nsim in 1 : Nsim){
  #simulate dataset $\mathbb{D}$
  x_train = runif(n_train, xmin, xmax)
  delta_train = rnorm(n_train, 0, sqrt(sigsq)) #mean zero, variance sigsq always (not dependent on value of x)
  y_train = x_train^2 + delta_train
  
  #fit a model g | x's, delta's and save it
  g_model = lm(y_train ~ poly(x, 5, raw = TRUE), data.frame(x = x_train))
  training_gs[nsim, ] = coef(g_model)
  
  #generate oos dataset and save residuals on oos data
  x_test = runif(n_test, xmin, xmax)
  delta_test = rnorm(n_test, 0, sigsq) #mean zero, variance sigsq always (not dependent on value of x)
  y_test = x_test^2 + delta_test
  y_hat_test = predict(g_model, data.frame(x = x_test))
  all_residuals[nsim, ] = y_test - y_hat_test
}
```

Take a look at the mse:

```{r}
mse = mean(c(all_residuals)^2)
mse
```

Much higher! Why? Variance went up!

Let's visualize the bias

```{r}
g_average = colMeans(training_gs)
f = function(x){x^2}
x = seq(xmin, xmax, length.out = resolution)
ggplot(data.frame(x = x, f = f(x)), aes(x, f)) + 
  geom_line(col = "darkgreen") + 
  stat_function(fun = function(x){g_average[1] + g_average[2] * x + g_average[3] * x^2 + g_average[4] * x^3 + g_average[5] * x^4 + g_average[6] * x^5}, col = "red")
```

Not much! Now acutllay compute the average bias squared of $g$:

```{r}
x = seq(xmin, xmax, length.out = resolution)
g_avg_x = g_average[1] + g_average[2] * x + g_average[3] * x^2 + g_average[4] * x^3 + g_average[5] * x^4 + g_average[6] * x^5
f_x = x^2
biases = f_x - g_avg_x
expe_bias_g_sq = mean(biases^2)
expe_bias_g_sq
```

This appears to have increased over last time ... but it's only because we're not running the regression infinite times. Remember this "expectation" is only an average.

What is the variance? Let's look at all lines:

```{r}
plot_obj = ggplot(data.frame(x = x)) + 
  xlim(xmin, xmax) + ylim(xmin^2, xmax^2)

for (nsim in 1 : min(Nsim, 50)){ #otherwise takes too long
  plot_obj = plot_obj + geom_line(data = data.frame(x = x, y = training_gs[nsim, 1] + training_gs[nsim, 2] * x + training_gs[nsim, 3] * x^2 + training_gs[nsim, 4] * x^3 + training_gs[nsim, 5] * x^4 + training_gs[nsim, 6] * x^5), mapping = aes(x, y), col = "blue")
}

plot_obj +
  stat_function(fun = function(x){g_average[1] + g_average[2] * x + g_average[3] * x^2 + g_average[4] * x^3 + g_average[5] * x^4 + g_average[6] * x^5}, col = "red", lwd = 2)
```

It looks awful!!!

Now actually compute the average variance numerically:

```{r}
x = seq(xmin, xmax, length.out = resolution)

expe_g_x = g_average[1] + g_average[2] * x + g_average[3] * x^2 + g_average[4] * x^3 + g_average[5] * x^4 + g_average[6] * x^5

var_x_s = array(NA, Nsim)
for (nsim in 1 : Nsim){
  g_x = training_gs[nsim, 1] + training_gs[nsim, 2] * x + training_gs[nsim, 3] * x^2 + training_gs[nsim, 4] * x^3 + training_gs[nsim, 5] * x^4 + training_gs[nsim, 6] * x^5
  var_x_s[nsim] = mean((g_x - expe_g_x)^2)
}

expe_var_g = mean(var_x_s)
expe_var_g
```

Any more complexity than you need allows for overfitting!

Now check the equivalence

```{r}
mse
sigsq
expe_bias_g_sq
expe_var_g
sigsq + expe_bias_g_sq + expe_var_g
```

# Bias - Variance Decomposition of MSE in Regression Trees

Let's return to the simulated sine curve data which we used to introduce regression trees.

```{r}
rm(list = ls())
n_train = 100
n_test = 500
xmin = 0
xmax = 10
sigsq = 0.09
Nsim = 250
resolution = 1000
```

And load the tree package:

```{r}
options(java.parameters = "-Xmx4000m")
# pacman::p_install_gh("kapelner/YARF", subdir = "YARF", ref = "dev")
pacman::p_load(YARF, tidyverse)
```

Now let's generate lots of different datasets and fit many tree models. Note there's a new argument `calculate_oob_error = FALSE`. This is here for speed only. Ignore this for now as we will go over what this means later in detail.

```{r}
training_gs = list() #storing entire objects - need a hash
all_residuals = matrix(NA, nrow = Nsim, ncol = n_test)

for (nsim in 1 : Nsim){
  #simulate dataset $\mathbb{D}$
  x_train = runif(n_train, xmin, xmax)
  delta_train = rnorm(n_train, 0, sqrt(sigsq))
  y_train = sin(x_train) + delta_train
  # ggplot(data.frame(x = x, y = y), aes(x, y)) + geom_point(lwd=0.6) 
  
  #fit a model g | x's, delta's and save it
  g_model = YARFCART(data.frame(x = x_train), y_train, calculate_oob_error = FALSE)
  training_gs[[nsim]] = g_model
  
  #generate oos dataset and save residuals on oos data
  x_test = runif(n_test, xmin, xmax)
  delta_test = rnorm(n_test, 0, sqrt(sigsq)) #mean zero, variance sigsq always (not dependent on value of x)
  y_test = sin(x_test) + delta_test
  y_hat_test = predict(g_model, data.frame(x = x_test))
  all_residuals[nsim, ] = y_test - y_hat_test
}
```

Let's look at the last data set to remind ourselves of the problem setting:

```{r}
ggplot(data.frame(x = x_train, y = y_train)) + geom_point(aes(x, y))
```


What does the storage of all the models look like?

```{r}
head(training_gs)
```

Take a look at the mse:

```{r}
mse = mean(c(all_residuals)^2)
mse
```

Let's visualize the bias

```{r}
#generate x and the truth
x = seq(xmin, xmax, length.out = resolution)
f_x = sin(x)

#now estimate the expectation of g by averaging all the different models
g_avg_x = array(0, resolution)
for (nsim in 1 : Nsim){
  g_nsim = training_gs[[nsim]]
  g_avg_x = g_avg_x + predict(g_nsim, data.frame(x = x))
}
g_avg_x = g_avg_x / Nsim #average of all models

#now plot
ggplot(data.frame(x = x, f = f_x, expe_g = g_avg_x)) + 
  geom_line(aes(x, f), col = "darkgreen") + 
  geom_line(aes(x, expe_g), col = "red")
```

Not much! Now actually compute the average bias squared of $g$:

```{r}
biases = f_x - g_avg_x
expe_bias_g_sq = mean(biases^2)
expe_bias_g_sq
```

This is small - why??

It's because trees are so expressive and have such model complexity that they can nail almost any true $f$ function!

That means the MSE save the irreducible noise is coming from the variance. Let's look at all lines:

```{r}
plot_obj = ggplot(data.frame(x = x)) + 
  xlim(xmin, xmax) #+ ylim(xmin^2, xmax^2)

for (nsim in 1 : min(Nsim, 100)){ #otherwise takes too long
  g_nsim = training_gs[[nsim]]
  g_x = predict(g_nsim, data.frame(x = x))
  plot_obj = plot_obj + geom_line(data = data.frame(x = x, y = g_x), aes(x = x, y = y), col = "blue")
}

plot_obj +
  geom_line(data = data.frame(x = x, expe_g = g_avg_x), mapping = aes(x, expe_g), col = "red", lwd = 1.5)
```

It looks awful!!!

Now actually compute the average variance numerically:

```{r}
x = seq(xmin, xmax, length.out = resolution)

var_x_s = array(NA, min(Nsim, 50))
for (nsim in 1 : min(Nsim, 50)){ #otherwise takes too long
  g_nsim = training_gs[[nsim]]
  g_x = predict(g_nsim, data.frame(x = x))
  var_x_s[nsim] = mean((g_x - g_avg_x)^2)
}

expe_var_g = mean(var_x_s)
expe_var_g
```

Any more complexity than you need allows for overfitting!

Now check the equivalence

```{r}
mse
sigsq
expe_bias_g_sq
expe_var_g
sigsq + expe_bias_g_sq + expe_var_g
```

# Model Averaging via Bootstrap Aggregation ("Bagging")

```{r}
rm(list = ls())
n_train = 100
n_test = 500
xmin = 0
xmax = 10
sigsq = 0.09
num_trees = 250
resolution = 1000
```

And load the tree package:

```{r}
options(java.parameters = "-Xmx4000m")
# pacman::p_install_gh("kapelner/YARF", subdir = "YARF", ref = "dev")
pacman::p_load(YARF)
```

Now let's generate one dataset (only ONE) 

```{r}
x_train = runif(n_train, xmin, xmax)
delta_train = rnorm(n_train, 0, sqrt(sigsq))
y_train = sin(x_train) + delta_train
ggplot(data.frame(x = x_train, y = y_train)) + geom_point(aes(x, y))
```

Spend a moment to appreciate that this is all we're going to have. There is a lot of irreducible noise that is going to prevent us from finding the sine curve.

Now we create many bootstrap samples to create similiar but different datasets and fit many tree models:

```{r}
bootstrapped_gs = list() #storing entire objects - need a hash

for (t in 1 : num_trees){
  #fit a model g | x's, delta's and save it
  bootstrap_indices_t = sample(1 : n_train, replace = TRUE)
  g_model = YARFCART(data.frame(x = x_train[bootstrap_indices_t]), y_train[bootstrap_indices_t], calculate_oob_error = FALSE)
  bootstrapped_gs[[t]] = g_model
}
```

Now let's aggregate all models constructed with bootstrap samples together by averaging and see what it looks like relative to real $f$:

```{r}
#generate x and the truth
x = seq(xmin, xmax, length.out = resolution)
f_x = sin(x)

#create the bagged model predictions
g_bagged = array(0, resolution)
for (t in 1 : num_trees){
  g_t = bootstrapped_gs[[t]]
  g_bagged = g_bagged + predict(g_t, data.frame(x = x))
}
g_bagged = g_bagged / num_trees #average of all models

#now plot
ggplot(data.frame(x = x, f = f_x, g_bagged = g_bagged)) + 
  geom_line(aes(x, f), col = "darkgreen") + 
  geom_line(aes(x, g_bagged), col = "red")
```
  
That's pretty good considering the plot of the raw data from above.

Now let's see if this is truly better than one tree. The way to do this is generate lots of datasets and try to estimate mse (with bias and variance).

To do this, we employ a convenience method called `YARFBAG`. We first define a number of trees in the "bag":

```{r}
num_trees = 250 #called "M" in class
```

And then we run the same simulation as we did above with one tree.

```{r}
Nsim = 250
training_gs = list() #storing entire objects - need a hash
all_residuals = matrix(NA, nrow = Nsim, ncol = n_test)

for (nsim in 1 : Nsim){
  #simulate dataset $\mathbb{D}$
  x_train = runif(n_train, xmin, xmax)
  delta_train = rnorm(n_train, 0, sqrt(sigsq))
  y_train = sin(x_train) + delta_train
  # ggplot(data.frame(x = x, y = y), aes(x, y)) + geom_point(lwd=0.6) 
  
  #fit a model g | x's, delta's and save it
  g_model = YARFBAG(data.frame(x = x_train), y_train, num_trees = num_trees, calculate_oob_error = FALSE)
  training_gs[[nsim]] = g_model
  
  #generate oos dataset and save residuals on oos data
  x_test = runif(n_test, xmin, xmax)
  delta_test = rnorm(n_test, 0, sqrt(sigsq)) #mean zero, variance sigsq always (not dependent on value of x)
  y_test = sin(x_test) + delta_test
  y_hat_test = predict(g_model, data.frame(x = x_test))
  all_residuals[nsim, ] = y_test - y_hat_test
}
```

Let's look at the last data set to remind ourselves of the problem setting:

```{r}
ggplot(data.frame(x = x_train, y = y_train)) + geom_point(aes(x, y))
```


What does the storage of all the models look like?

```{r}
head(training_gs)
```

Take a look at the mse:

```{r}
mse = mean(c(all_residuals)^2)
mse
```

Let's visualize the bias

```{r}
#generate x and the truth
x = seq(xmin, xmax, length.out = resolution)
f_x = sin(x)

#now estimate the expectation of g by averaging all the different models
g_avg_x = array(0, resolution)
for (nsim in 1 : Nsim){
  g_nsim = training_gs[[nsim]]
  g_avg_x = g_avg_x + predict(g_nsim, data.frame(x = x))
}
g_avg_x = g_avg_x / Nsim #average of all models

#now plot
ggplot(data.frame(x = x, f = f_x, expe_g = g_avg_x)) + 
  geom_line(aes(x, f), col = "darkgreen") + 
  geom_line(aes(x, expe_g), col = "red")
```

Not much! If anything it should be *better* than the bias of a single tree!

Now actually compute the average bias squared of $g$:

```{r}
biases = f_x - g_avg_x
expe_bias_g_sq = mean(biases^2)
expe_bias_g_sq
```

Almost zilch.

Now... variance?

```{r}
plot_obj = ggplot(data.frame(x = x)) + 
  xlim(xmin, xmax) #+ ylim(xmin^2, xmax^2)

for (nsim in 1 : min(Nsim, 50)){ #otherwise takes too long
  g_nsim = training_gs[[nsim]]
  g_x = predict(g_nsim, data.frame(x = x))
  plot_obj = plot_obj + geom_line(data = data.frame(x = x, y = g_x), aes(x = x, y = y), col = "blue")
}

plot_obj +
  geom_line(data = data.frame(x = x, expe_g = g_avg_x), mapping = aes(x, expe_g), col = "red", lwd = 1.5)
```

There is variance, but less than previously when only one tree was used.

Now actually compute the average variance numerically:

```{r}
x = seq(xmin, xmax, length.out = resolution)

var_x_s = array(NA, min(Nsim, 50))
for (nsim in 1 : min(Nsim, 50)){ #otherwise takes too long
  g_nsim = training_gs[[nsim]]
  g_x = predict(g_nsim, data.frame(x = x))
  var_x_s[nsim] = mean((g_x - g_avg_x)^2)
}

expe_var_g = mean(var_x_s)
expe_var_g
```

Any more complexity than you need allows for overfitting!

Now check the equivalence

```{r}
mse
sigsq
expe_bias_g_sq
expe_var_g
sigsq + expe_bias_g_sq + expe_var_g
```

We have a better algorithm!

# Bagged Trees vs. a Linear Model

First we'll load the packages and data:

```{r}
rm(list = ls())
options(java.parameters = "-Xmx8000m")
# pacman::p_install_gh("kapelner/YARF", subdir = "YARF", ref = "dev")
# pacman::p_load(YARF, ggplot2)
library(YARF)
diamonds_sample = diamonds[sample(1 : nrow(diamonds), 2000), ]
boston = MASS::Boston
cancer = MASS::biopsy
cancer$ID = NULL
cancer = na.omit(cancer)
pacman::p_load_gh("coatless/ucidata")
data(adult)
adult = na.omit(adult) #kill any observations with missingness
adult_sample = adult[sample(1 : nrow(adult), 2000), ]
```

Let's look at the boston housing data first

```{r}
prop_test = 0.1
test_indices = sample(1 : nrow(boston), round(prop_test * nrow(boston)))
boston_test = boston[test_indices, ]
y_test = boston_test$medv
X_test = boston_test
X_test$medv = NULL
train_indices = setdiff(1 : nrow(boston), test_indices)
boston_train = boston[train_indices, ]
y_train = boston_train$medv
X_train = boston_train
X_train$medv = NULL

mod_lin = lm(y_train ~ ., X_train)
y_hat_test_lin = predict(mod_lin, X_test)
s_e_lin = sd(y_test - y_hat_test_lin)
s_e_lin

num_trees = 500
mod_bag = YARFBAG(X_train, y_train, num_trees = num_trees, calculate_oob_error = FALSE)
y_hat_test_bag = predict(mod_bag, X_test)
s_e_bag = sd(y_test - y_hat_test_bag)
s_e_bag

cat("oos standard error reduction:", (1 - s_e_bag / s_e_lin) * 100, "%\n")
```

How much room was there to improve?

```{r}
summary(mod_lin)$r.squared
```

Whole lot of room to improve! That extra 26% or so of unexplained variance is split between bias and irreducible error due to unknown information. The bagged trees can get rid of the bias and minimize variance while doing so. And it did a great job!

Now the diamonds data:

```{r}
n_train = 500

training_indices = sample(1 : nrow(diamonds), n_train)
diamonds_train = diamonds[training_indices, ]
y_train = diamonds_train$price
X_train = diamonds_train
X_train$price = NULL


test_indices = setdiff(1 : nrow(diamonds), training_indices)
diamonds_test = diamonds[test_indices, ]
y_test = diamonds_test$price
X_test = diamonds_test
X_test$price = NULL


mod_lin = lm(y_train ~ ., X_train)
y_hat_test_lin = predict(mod_lin, X_test)
s_e_lin = sd(y_test - y_hat_test_lin)
s_e_lin

mod_bag = YARFBAG(X_train, y_train, calculate_oob_error = FALSE)
y_hat_test_bag = predict(mod_bag, X_test)
s_e_bag = sd(y_test - y_hat_test_bag)
s_e_bag

cat("oos standard error reduction:", (1 - s_e_bag / s_e_lin) * 100, "%\n")
```

```{r}
summary(mod_lin)$r.squared
```

Not a whole lot of room to improve! That extra 8% of unexplained variance is split between bias and irreducible error due to unknown information. But it did the best it can. It is likely what's remaining is due to information we are not privy to.


# Validation in Bagging?

We are using the "bootstrap" to get the trees. Can we do model validation in the same step? 

The answer is yes. For every tree, there was a bootstrap sample of the training set used to build the tree. But there are observations in $\mathbb{D}$ that are not in the bootstrap sample! About 1/3 on average are left out i.e. "out of bag (oob)". Over many trees, there are different oob subsets than become the full data set. So you actually have validation in a way on the whole dataset kind of like K-fold cross validation. Supposedly this validation is similar to K=2 in terms of performance. It is what everyone seems to use. 

Let's load the data and packages from last class and plot the data:

```{r}
rm(list = ls())
n_train = 100
n_test = 500
xmin = 0
xmax = 10
sigsq = 0.09
num_trees = 500
x_train = runif(n_train, xmin, xmax)
delta_train = rnorm(n_train, 0, sqrt(sigsq))
y_train = sin(x_train) + delta_train
ggplot(data.frame(x = x_train, y = y_train)) + geom_point(aes(x, y))
```

Let's look at one bagged tree model and compute OOB errors after construction:

```{r}
bagged_tree_mod = YARFBAG(data.frame(x = x_train), y_train, num_trees = num_trees) #TRUE is the default! Why? It's something you want to know!
bagged_tree_mod
```

How did this work? Let's look at the oob sets:

```{r}
cat("bootstrap indices:")
sort(bagged_tree_mod$bootstrap_indices[[1]])
cat("oob:")
sort(setdiff(1 : n_train, bagged_tree_mod$bootstrap_indices[[1]]))
cat("bootstrap indices:")
bagged_tree_mod$bootstrap_indices[[2]]
cat("oob:")
setdiff(1 : n_train, bagged_tree_mod$bootstrap_indices[[2]])
```

It took predictions on each tree on the oob set, averaged by observation across trees and then averaged across observation averages.


# Random Forests


What do we have now? We have model selection is done within training. And training and validation are done in a single step! No more costly K-fold CV with 3 splits!

But can it get any better? YES. As you saw, the variance terms can be shrunk further the more decorrelated the trees become. We do this now by introducing randomness into the splits by choosing only a subset of the features to split on randomly. The trees are then grown as normal. Then the we model average many trees via bagging. And that's random forests!

Quick demo with the boston housing data and oob validation:

```{r}
rm(list = ls())
boston = MASS::Boston
y = boston$medv
X = boston
X$medv = NULL

mod_bag = YARFBAG(X, y, num_trees = 500)
mod_bag
mod_rf = YARF(X, y, num_trees = 500)
mod_rf
```

Gain in decorrelation?

```{r}
cat("gain: ", (mod_rf$pseudo_rsq_oob - mod_bag$pseudo_rsq_oob) / mod_bag$pseudo_rsq_oob * 100, "%\n")
```

For this example, not much. How about on the diamonds dataset?

```{r}
diamonds = ggplot2::diamonds

rm(list = ls())
num_trees = 500
n_train = 500

training_indices = sample(1 : nrow(diamonds), n_train)
diamonds_train = diamonds[training_indices, ]
y_train = diamonds_train$price
X_train = diamonds_train
X_train$price = NULL

mod_bag = YARFBAG(X_train, y_train, num_trees = num_trees, calculate_oob_error = FALSE)
mod_rf = YARF(X_train, y_train, num_trees = num_trees, calculate_oob_error = FALSE)

test_indices = setdiff(1 : nrow(diamonds), training_indices)
diamonds_test = diamonds[test_indices, ]
y_test = diamonds_test$price
X_test = diamonds_test
X_test$price = NULL

mse_bag = sum((y_test = predict(mod_bag, X_test))^2) / nrow(X_test)
mse_rf = sum((y_test = predict(mod_rf, X_test))^2) / nrow(X_test)
cat("gain: ", (mse_bag - mse_rf) / mse_bag * 100, "%\n")
```

Also, not much, but real.

