---
title: "Practice Lecture 13 MATH 390.4 Queens College"
author: "Professor Adam Kapelner"
date: "May 7, 2019"
---

# Joins

```{r}
pacman::p_load(nycflights13, tidyverse, magrittr)
data(weather)
str(weather)
data(airports)
str(airports)
```

Note how the weather and airports datasets contain a common feature: name of airport. It is called `FAA` in airports and `origin` in weather.

First we rename the column in weather to match the column in airports:

```{r}
weather %<>% rename(faa = origin)
```

We also pare down the datasets so we can see the joins more clearly:

```{r}
airports %<>% select(faa, lat, lon)
weather %<>% select(faa, time_hour, temp, humid, wind_speed, pressure, wind_gust)
```

Some features just aren't measured that often e.g. `wind_gust`.

Let's do some joins. 

```{r}
head(airports)
head(weather)
airports_and_weather = left_join(airports, weather, by = "faa")
airports_and_weather %>% sample_n(500)
```


```{r}
head(airports)
head(weather)
airports_and_weather = right_join(airports, weather, by = "faa")
airports_and_weather %>% sample_n(500)
```

Just an example... drop `EWR` from airports:

```{r}
airports_without_EWR = airports %>%
  filter(faa != "EWR")
right_join(airports_without_EWR, weather, by = "faa") %>% sample_n(500)
```

```{r}
head(airports)
head(weather)
airports_and_weather = inner_join(airports, weather, by = "faa")
airports_and_weather %>% sample_n(500)
```




```{r}
head(airports)
head(weather)
airports_and_weather = full_join(airports, weather, by = "faa")
airports_and_weather %>% sample_n(500) %>% arrange(faa)
```


```{r}
airports_without_EWR = airports %>%
  filter(faa != "EWR")
full_join(airports_without_EWR, weather, by = "faa") %>% sample_n(500)
```

There is also `semi_join` and `anti_join` that do the opposite of joining. In my experience, these use cases are limited.

# Correlation does not Imply Causation

Take a look at the following real data:

```{r}
spurious = data.frame(
  yearly_divorce_rate_maine_per_1000 = c(5,4.7,4.6,4.4,4.3,4.1,4.2,4.2,4.2,4.1),
  yearly_US_consumption_margarine_per_capita = c(8.2,7,6.5,5.3,5.2,4,4.6,4.5,4.2,3.7)
)

ggplot(spurious, aes(x = yearly_divorce_rate_maine_per_1000, y = yearly_US_consumption_margarine_per_capita)) +
  geom_point() + geom_smooth()
```

How did this happen? 

Well, we can imagine doing the same thing.


```{r}
y = spurious$yearly_US_consumption_margarine_per_capita
n = length(y)

r = 100000
best_abs_corr = 0
best_random_xs = NULL
for (i in 1 : r){
  x = rnorm(n)
  
  random_abs_corr = abs(cor(x, y))
  if (random_abs_corr > best_abs_corr){
    best_abs_corr = random_abs_corr
    best_random_xs = x
  }
}
spurious$best_random_xs = best_random_xs
ggplot(spurious, aes(x = best_random_xs, y = yearly_US_consumption_margarine_per_capita)) +
  geom_point() + geom_smooth() + ggtitle(paste("Spurious Correlation has |r| = ", round(best_abs_corr, 3)))
```

So what's the narrative here? If you look through a gajillion random features that have no causal connection with the phenomenon $y$, you will eventually find something that "clicks". However, this will be vanish if you keep collecting data. Anything that is built upon falsehood will crumble!

#An Example of Correlation without Causation

When does correlation really not imply causation? From class, we spoke about the phenomenon $y$ = "num car accidents" with observed feature "x" = "num umbrellas sold" but common cause $z$ = "rain amount". It is clear the umbrella sales has *no causal* relationship with car accidents. But they *are correlated* because they are linked by a common cause. Here's the data example that makes this clear.

The data generating process as specified by the causal diagram looks as follows:

```{r}
rm(list = ls())
set.seed(1)
n = 300
sigma = 0.3

umbrella_example_data = data.frame(
  z_rainfall = runif(n, 0, 6) #here's the common cause - rainfall
)
umbrella_example_data$x_umbrella_sales = umbrella_example_data$z_rainfall^2 + rnorm(n, sigma) #x is a variable that is driven by z with noise
umbrella_example_data$y_car_accidents = umbrella_example_data$z_rainfall + rnorm(n, sigma) #y is a variable driven by z with noise
```

So we only see $x$ and $y$. Here's what it appears as:

```{r}
pacman::p_load(ggplot2)
ggplot(umbrella_example_data) +
  aes(x = x_umbrella_sales, y = y_car_accidents) +
  geom_point() + 
  geom_smooth(method = "lm")
```

and the model looks like:

```{r}
mod = lm(y_car_accidents ~ x_umbrella_sales, umbrella_example_data)
round(coef(mod), 3)
```

So what's the interpretation of the coefficient for $x$? ...

What you can't say is that $x$ is a causal contributor to $y$! You may want to say it, but you can't!

Now let's build a model of $y$ linear in both $x$ and $z$. What happens?

```{r}
mod = lm(y_car_accidents ~ x_umbrella_sales + z_rainfall, umbrella_example_data)
round(coef(mod), 3)
```

The effect of $x$ is gone!! Why? If you keep $z$ constant, the sole true causal factor in $y$, manipulating $x$ won't matter anymore!

Why is this? Well, you can look at how x affects y in local areas of z for instance.

```{r}
b = 0.8; a = 0.7
z_small_indices = umbrella_example_data$z_rainfall < 
  quantile(umbrella_example_data$z_rainfall, b) &
  umbrella_example_data$z_rainfall >
  quantile(umbrella_example_data$z_rainfall, a)

ggplot(umbrella_example_data[z_small_indices, ]) +
  aes(x = x_umbrella_sales, y = y_car_accidents) +
  geom_point()
```

If you force the common cause (lurking variable) to be an approximate constant, then you won't see any affect of x on y.


#Classification Trees and Confusion Tables

Let's load up the adult dataset where the response is 1 if the person makes more than $50K per year and 0 if they make less than $50K per year.

```{r}
pacman::p_load_gh("coatless/ucidata")
data(adult)
adult %<>% 
  na.omit #kill any observations with missingness
str(adult)
?adult
```

Let's use samples of 2,000 to run experiments:

```{r}
test_size = 2000
train_indices = sample(1 : nrow(adult), test_size)
adult_train = adult[train_indices, ]
y_train = adult_train$income
X_train = adult_train
X_train$income = NULL
test_indices = sample(setdiff(1 : nrow(adult), train_indices), test_size)
adult_test = adult[test_indices, ]
y_test = adult_test$income
X_test = adult_test
X_test$income = NULL
```

Load the tree package:

```{r}
options(java.parameters = "-Xmx4000m")
# pacman::p_install_gh("kapelner/YARF", subdir = "YARF", ref = "dev")
pacman::p_load(YARF)
```

Make a tree and compute in-sample and out of sample fits:

```{r}
tree_mod = YARFCART(X_train, y_train)
y_hat_train = predict(tree_mod, X_train)
y_hat_test = predict(tree_mod, X_test)
```

Let's look at the confusion table in-sample:

```{r}
table(y_train, y_hat_train)
```

There are no errors here! Thus, precision and recall are both 100%. This makes sense because classification trees overfit.

Let's do the same oos:

```{r}
oos_conf_table = table(y_test, y_hat_test)
oos_conf_table
```

We didn't do as well (of course). Let's calculate some performance metrics. We assume ">50k" is the "positive" category and "<=50k" is the "negative" category. Note that this choice is arbitrary and everything would just be switched if we did it the other way.

```{r}
n = sum(oos_conf_table)
n
fp = oos_conf_table[1, 2]
fn = oos_conf_table[2, 1]
tp = oos_conf_table[2, 2]
tn = oos_conf_table[1, 1]
num_pred_pos = sum(oos_conf_table[, 2])
num_pred_neg = sum(oos_conf_table[, 1])
num_pos = sum(oos_conf_table[2, ])
num_neg = sum(oos_conf_table[1, ])
acc = (tp + tn) / n
acc
misclassifcation_error = 1 - acc
misclassifcation_error
precision = tp / num_pred_pos
precision
recall = tp / num_pos
recall
false_discovery_rate = 1 - precision
false_discovery_rate
false_omission_rate = fn / num_pred_neg
false_omission_rate
```

Let's see how this works on a dataset whose goal is classification for more than 2 levels. Note: this is only possible now with trees!

```{r}
rm(list = ls())
pacman::p_load(mlbench)
data(LetterRecognition)
LetterRecognition = na.omit(LetterRecognition) #kill any observations with missingness
str(LetterRecognition)
?LetterRecognition
```

Now we split the data:

```{r}
test_samp = 2000
train_indices = sample(1 : nrow(LetterRecognition), test_samp)
ltr_train = LetterRecognition[train_indices, ]
y_train = ltr_train$lettr
X_train = ltr_train
X_train$lettr = NULL
test_indices = sample(setdiff(1 : nrow(LetterRecognition), train_indices), test_samp)
ltr_test = LetterRecognition[test_indices, ]
y_test = ltr_test$lettr
X_test = ltr_test
X_test$lettr = NULL
```

And fit a tree model and its in-sample and oos fits:

```{r}
tree_mod = YARFCART(X_train, y_train)
y_hat_train = predict(tree_mod, X_train)
y_hat_test = predict(tree_mod, X_test)
```

Take a look at the in-sample confusion matrix:

```{r}
table(y_train, y_hat_train)
```

Perfecto... as expected... 

Now the oos confusion matrix:

```{r}
oos_confusion_table = table(y_test, y_hat_test)
oos_confusion_table
```

Hard to read. Let's make it easier to read by blanking out the diagonal and looking at entried only >= 5:

```{r}
oos_confusion_table[oos_confusion_table < 5] = ""
diag(oos_confusion_table) = "."
oos_confusion_table
```

What's it using to determine letter?

```{r}
illustrate_trees(tree_mod, max_depth = 3, open_file = TRUE)
```

Where did these features comes from?? Deep learning helps to create the features from the raw pixel data. Wish I had a whole next semester to discuss this...



# Missingness

Take a look at the weather dataset again:

```{r}
pacman::p_load(nycflights13, tidyverse, magrittr)
data(weather)
str(weather)
```


