---
title: "Lab 8"
author: "Your Name Here"
output: pdf_document
date: "11:59PM April 7, 2019"
---

Load the `ggplot2` library and its dataset called `mpg`. Print out a summary of the dataset using `summary` and `str`. 

```{r}
pacman::p_load(ggplot2)
data(mpg)
mpg$drv = factor(mpg$drv)
mpg$manufacturer = factor(mpg$manufacturer)
mpg$model = factor(mpg$model)
mpg$fl = factor(mpg$fl)
mpg$class = factor(mpg$class)
mpg$cyl = factor(mpg$cyl)
summary(mpg)
?mpg
str(mpg)
```

Visualize a histogram then a density estimate of the `displ` variable, the engine displacement. Use `labs` to create a `title`, `subtitle`, `caption` and x-label via `x` and y-label via `y`. Do this for every single illustration in this lab. 

```{r}
ggplot(mpg) + 
  aes(displ) +
  geom_histogram() +
  labs(title = "Engine Displacement Histogram" , subtitle = "", x = "Displacement (L)", caption = "Source: EPA 2008 Fuel Economy Dataset")


ggplot(mpg) + 
  aes(displ) +
  geom_density(fill = "blue", alpha = 0.2, col = "white") +
  labs(title = "Engine Displacement Density" , subtitle = "", x = "Displacement (L)", caption = "Source: EPA 2008 Fuel Economy Dataset")

```


Visualize a histogram the `displ` variable, but then fill the color of the bar by the `class` of the car. You will have to pass `class` in as the `fill` in the aesthetic of the histogram. 

```{r}
ggplot(mpg) + 
  aes(displ) +
  geom_histogram(aes(fill = class)) +
  labs(title = "Engine Displacement Histogram" , subtitle = "Colored by Vehicle Type", x = "Displacement (L)", caption = "Source: EPA 2008 Fuel Economy Dataset")


```

Visualize overlapping densities of `cty` (city miles per gallon) and `hwy` (highway miles per gallon) using two colors with an alpha blend.  

```{r}
ggplot(mpg) +
  geom_density(aes(cty), fill = "red", col = "white", alpha = 0.2) +
  geom_density(aes(hwy), fill = "blue", col = "white", alpha = 0.2) +
  xlim(5, 50) +
  labs(title = "Fuel Efficiency Density" , subtitle = "City in Red and Highway in Blue", x = "miles per gallon", caption = "Source: EPA 2008 Fuel Economy Dataset")
```

Plot `cty` (city miles per gallon) vs `hwy` (highway miles per gallon) and draw a best fit line with a confidence region of that line.

```{r}
ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  labs(title = "City vs Highway Fuel Efficiency", subtitle = "With best fit line and confidence interval", x = "City (mpg)", y = "Highway (mpg)", caption = "Source: EPA 2008 Fuel Economy Dataset")
```

Plot `cty` (city miles per gallon) vs `hwy` (highway miles per gallon) and draw a best fit non-parametric functional relationship with a confidence region of that relationship.

```{r}
ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point() +
  geom_smooth() + 
  labs(title = "City vs Highway Fuel Efficiency", subtitle = "With best fit line and confidence interval", x = "City (mpg)", y = "Highway (mpg)", caption = "Source: EPA 2008 Fuel Economy Dataset")
```


Plot `cty` (city miles per gallon) vs `hwy` (highway miles per gallon) and then try to visualize *as many other variables* as you can visualize effectively on the same plot. Try text, color, size, shape, etc.

```{r}
ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point(aes(col = drv, shape = cyl, size = displ)) +
  geom_text(aes(label = fl)) +
  labs(title = "City vs Highway Fuel Efficiency", x = "City (mpg)", y = "Highway (mpg)", caption = "Source: EPA 2008 Fuel Economy Dataset")
```


Convert `cyl` to an ordinal factor. Then use the package `dplyr` to retain only cars with 4, 6, 8 cylinders in the dataset. Then make a canonical illustration of `cty` by `cyl`.

```{r}
mpg$cyl = factor(mpg$cyl, ordered = TRUE)
#mpg$cyl
pacman::p_load(dplyr)
mpg = mpg %>%
  filter(cyl %in% c(4, 6, 8))
ggplot(mpg, aes(x = cyl, y = cty)) +
  geom_boxplot() +
  labs(title = "City Fuel Efficiency by Number of Cylnders", x = "Number of Cylnders", y = "City (mpg)", caption = "Source: EPA 2008 Fuel Economy Dataset")
```


Load the `stringr` library. Use the `str_detect` function in this libary to rewrite the `trans` variable in the data frame to be just "manual" or "automatic".

```{r}
#TO-DO
```

Now visualize `cty` by `trans` via two overlapping alpha-blended densities. 

```{r}
#TO-DO
```

Now visualize `cty` by `trans` via a box and whisker plot.

```{r}
#TO-DO
```


Now visualize `cty` by `drv` by `trans` via two box and whisker plots horizontally laid out.

```{r}
#TO-DO
```

Now visualize `cty` by `class` via a violin plot. Look at the ggplot cheatsheet!

```{r}
ggplot(mpg, aes(x = class, y = cty)) +
  geom_violin(fill = "brown", alpha = 0.4) +
  labs(title = "City Fuel Efficiency by Vehicle Type", x = "Class", y = "City (mpg)", caption = "Source: EPA 2008 Fuel Economy Dataset")
```

Make a pie chart of `class`.

Visualize `trans` vs `class`. Look at the ggplot cheatsheet!

```{r}
#TO-DO
```

Using the package `ggExtra`'s `ggMarginal` function, look at the `hwy` by `year` and plot the marginal density on both the x and y axes.

```{r}
#TO-DO
```

Using the package `ggcorrplot`'s `ggcorrplot` function, look at the correlations for all variables in this dataset that are legal in a corrrelogram. Use dplyr to `select_if` the variable is appropriate.

```{r}
#TO-DO
```

Use the `stat_ecdf` function to plot the estimated cumulative distribution of `cty'.

```{r}
#TO-DO
```

Create a data generating process where $x$ is uniform between 0 and 10 and $y$ is $x^2$ plus $N(0,1)$ noise. Plot $n = 200$ points and then plot the quadratic relationship $y = x^2$ using the function `stat_function`.

```{r}
#TO-DO
```

We now move to Rcpp. Load the library.

```{r}
#TO-DO
```

Write an R function `is_odd` and a C++ function `is_odd_cpp` that evaluates if a number is odd and returns true if so.


```{r}
#TO-DO
```

Using `system.time', run both functions 1,000,000 times on the numbers 1, 2, ..., 1000000. Who is faster and by how much?

```{r}
#TO-DO
```

Write an R function `fun` and a C++ function `fun_cpp` that takes a natural number returns $n$ if $n$ is 0 or 1 otherwise the result of the function on $n - 1$ and $n - 2$. This is the function that returns the $n$th Fibonacci number.

```{r}
#TO-DO
```

Using `system.time', run both functions on the numbers 1, 2, ..., 100. Who is faster and by how much?

```{r}
#TO-DO
```

Write an R function `logs` and a C++ function `logs_cpp` that takes a natural number $n$ and returns an array of $ln(1), ln(2), ..., ln(n)$. 

```{r}
#TO-DO
```

Using `system.time', run both functions on the numbers 1, 2, ..., 1000000. Who is faster and by how much?

```{r}
#TO-DO
```

Write an R function `max_distances` and a C++ function `max_distances_cpp` that takes an $n \times p$ matrix $X$ and returns an $n \times n$ matrix called $D$ of NA's where the upper triangular portion above the diagonal is the max distances between the elements of the $i, j$th rows of $X$.


```{r}
#TO-DO
```

Create a matrix $X$ of $n = 1000$ and $p = 20$ filled with iid $N(0, 1)$ realizations. Using `system.time', calculate $D$ using both functions. Who is faster and by how much?

```{r}
#TO-DO
```





