---
title: "Lab 6"
author: "Your Name Here"
output: pdf_document
date: "11:59PM March 24, 2019"
---

Load the Boston Housing data and create the vector $y$ and the design matrix $X$.

```{r}
#TO-DO
```

Find the OLS estimate and OLS predictions without using `lm`.

```{r}
#TO-DO
```

Write a function spec'd as follows:

```{r}
#' Orthogonal Projection
#'
#' Projects vector a onto v.
#'
#' @param a   the vector to project
#' @param v   the vector projected onto
#'
#' @returns   a list of two vectors, the orthogonal projection parallel to v named a_parallel, 
#'            and the orthogonal error orthogonal to v called a_perpendicular
orthogonal_projection = function(a, v){
  
}
```


Try to project onto the column space of $X$ by projecting on each vector of $X$ individually and adding up the projections. You can use the function `orthogonal_projection`.

```{r}
#TO-DO
```

How much double counting occurred? Measure the magnitude relative to the true LS orthogonal projection.

```{r}
#TO-DO
```

Convert $X$ into $Q$ where $Q$ has the same column space as $X$ but has orthogonal columns. You can use the function `orthogonal_projection`. This is essentially gram-schmidt.

```{r}
#TO-DO
```

Make $Q$'s columns orthonormal.

```{r}
#TO-DO
```

Verify $Q^T$ is the inverse of $Q$.

```{r}
#TO-DO
```


Project $Y$ onto $Q$ and verify it is the same as the OLS fit.

```{r}
#TO-DO
```


Project $Y$ onto the columns of $Q$ one by one and verify it sums to be the projection onto the whole space.

```{r}
#TO-DO
```

Verify the OLS fit squared lengh is the sum of squared lengths of each of the orthogonal projections.

```{r}
#TO-DO
```

Rewrite the "The monotonicity of SSR" demo from the lec06 notes. Comment every line in detail. Write about what the plots means.

```{r}
#TO-DO
```

Rewrite the "Overfitting" demo from the lec06 notes. Comment every line in detail. Write about what the plots means.

```{r}
#TO-DO
```
