---
title: "Lab 5"
author: "Your Name Here"
output: pdf_document
date: "11:59PM March 16, 2019"
---

Load the Boston housing data frame and create the vector $y$ (the median value) and matrix $X$ (all other features) from the data frame.

```{r}
#TO-DO
```

Run the OLS linear model to get $b$, the vector of coefficients. Do not use `lm`.

```{r}
#TO-DO
```

Find the hat matrix for this regression.

```{r}
#TO-DO
```

Verify this is a projection matrix by verifying the two sufficient conditions. Use the `testthat` library.

```{r}
#TO-DO
```

Calculate $\hat{y}$ and $e$.

```{r}
#TO-DO
```

Calculate $R^2$ and RMSE.

```{r}
#TO-DO
```

Verify $\hat{y}$ and $e$ are orthogonal.

```{r}
#TO-DO
```

Verify $\hat{y} - \bar{y}$ and $e$ are orthogonal.

```{r}
#TO-DO
```

Find the cosine-squared of $y - \bar{y}$ and $\hat{y} - \bar{y}$ and verify it is the same as $R^2$.

```{r}
#TO-DO
```



```{r}
#TO-DO
```



```{r}
#TO-DO
```



```{r}
#TO-DO
```



```{r}
#TO-DO
```



```{r}
#TO-DO
```



```{r}
#TO-DO
```



```{r}
#TO-DO
```



```{r}
#TO-DO
```



```{r}
#TO-DO
```



```{r}
#TO-DO
```



