---
title: "Review Lab 1"
author: "Your Name Here"
output: pdf_document
---

These are exercises to practice for the exam. You do not need to hand this lab in as an assignment.

* Create a vector of 1000 draws from a normal distribution centered at 50 with variance 25.

```{r}
v = rnorm(1000, mean= 50, sd = sqrt(25))
v
```

* Round those distributions down to the closest integer.

```{r}
floor(v)
```

* Round those distributions up to the closest integer.

```{r}
ceiling(v)
```

* Create a vector that looks like 1,1,1,1,1,2,2,2,2,2, ..., 100,100,100,100,100. Do it in as few operations as possible.

```{r}
x = rep(1:100,each=5)
x
```

Use the `which` function to find the indices of numbers that are divisible by 7.

```{r}
which(x %% 7 == 0, arr.ind = TRUE)

```


* Create a nominal factor vector that looks like A,A,A,B,B,B,C,C,C,D,D,D,D, ..., Z,Z,Z,Z and make the reference category Z.

```{r}
x = rep(LETTERS,each=3)
x
```

* Create a list mapping "A" => 1, "B" => 2, ..., "Z" => 26

```{r}
map = list()

for(i in 1 : length(LETTERS)){
  map[[LETTERS[i]]] = i
}

map

x = 1:26
names(x) = LETTERS
x
as.list(x)

x = list()
x[LETTERS] = 1:26
x
```

* Use `lapply` to sum over the values in the list.

```{r}
?lapply

```


* Create a matrix of size 2 x 26 where the first row is the list's keys and the second row is the list's values.

```{r}
mat = matrix(c(names(map),as.numeric(map)) , ncol = 26,nrow = 2, byrow = TRUE)
mat

rbind(names(x), as.numeric(x))

```

* Write a script that runs a function f (assumed to return a scalar) on each column of a matrix X and aggregates the results in a vector.

```{r}
X = matrix(rnorm(100), nrow = 10)
f = sum
class(f)
v = rep(NA, ncol(X))

for (j in 1 : ncol(X)){
  v[j] = f(X[, j])
}

v


```

Load the boston housing data frame.

```{r}
data(Boston, package = "MASS")
```

Create a list object that holds 10 dataframes where each consists of the boston housing data frame whose observations are sampled with replacement.

```{r}
lst = list()
for(i in 1:10){
  lst[[i]] = Boston[sample(1 : nrow(Boston), replace = TRUE), ]
}
head(lst)
  
```


Cross-tab the variable chas with rad.


```{r}
?rad
?table

table(Boston$chas, Boston$rad)
```

Create a new variable in the data frame called `fake_y` equal to `3 * age + 6 * lstat - 100 * crim+`  iid random realizations from $N(0,1)$.

```{r}
Boston$fake_y = 3 * Boston$age + 6 * Boston$lstat - 100 * Boston$crim + rnorm(nrow(Boston))
summary(Boston$fake_y)
```

Find the deciles of `fake_y` i.e. the min, the 10th percentile, the 20th percentile, ..., the 90th percentile and the max.

```{r}
quantile(Boston$fake_y, probs = seq(0,1, .1))
```


Find the subset of the data frame where `fake_y` is negative.

```{r}
Boston[Boston$fake_y < 0, ]
```


Create a new variable in the data frame which binarizes `fake_y` by being 1 if positive and 0 if negative.

```{r}
Boston$fake_y_binary = ifelse(Boston$fake_y >= 0, 1, 0)
```
Fit an OLS model of 'fake_y` using `age` and find the estimates of the intercept and slope and RMSE and R^2.

```{r}
lmm = lm(fake_y ~ age, Boston)
coef(lmm)
summary(lmm)$r.squared
summary(lmm)$sigma
```


